---
title: "Benchmark Elevation Sources to assess gradient for active transportation"
output: github_document
bibliography: refs.bib
link-citations: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Different elevation sources may bring better or worse results, when computing slopes for a road network.  
In many cases, the resolution of raster digital elevation models (DEM) does not matter much, when we assume that they will be small enough for large segments - for instance, if we need to know the gradient of a 2km length highway, planned for motorized vehicles.  
But when physical effort matters, such as for active travel (walking, cycling), a 50m road with a 2% gradient might be very different that an 8% gradient.  
For smaller road segments, getting an accurate gradient value might be an issue, in particular when the free and open data sources do not provide a good resolution.  
In this example, we will see and compare the results of [slopes R package](https://github.com/ITSLeeds/slopes) for the same road network sample (in Lisbon, Portugal), using **four** different elevation data sources, with different resolution. 

## Literature

Google Maps Terrain is improving every year [@Harris2014; @Egypt2016; @Wang2017]. It started by using SRTM data, but now at some places, it recurs to LIDAR technology [@Scott2010], although not available yet for all places (see: https://en.wikipedia.org/wiki/National_lidar_dataset). It requires an API key for [elevation access](https://developers.google.com/maps/documentation/elevation/overview).

[OpenTopoData](https://www.opentopodata.org/#public-api ) provides a nice comparison of the available topographic data, regarding its resolution and coverage. Unfortunately, the open ones, with a good resolution, only cover USA and New Zeland. 


## Materials - Data sources
Open data sources.  

### For Elevation Models 

For this example, we will use and compare four elevation sources:

* NASA Digital Elevation Model, with 27m cell resolution
* MapBox-Terrain tiles, with 0.1 meter height increments (ref)
* Copernicus European DEM, with 25m cell resolution
* Instituto Superior Técnico Digital Elevation Model, with 10m cell resolution


##### NASA DEM
The SRTM NASA's mission 
Os dados do SRTM ([Shuttle Radar Topography Mission](https://www.usgs.gov/centers/eros/science/usgs-eros-archive-digital-elevation-shuttle-radar-topography-mission-srtm-1-arc)), uma missão da NASA, estão disponíveis gratuitamente, mas para uma resolução de ~30m, com erro da altimetria vertical de 16m. Para fazer donwload do tile correcto, pode-se também recorrer a um outro plugin do QGIS, o SRTM-Donwloader, e pedir para guardar o raster que cobre a shapefile da rede viária - é uma opção no QGIS, e é necessário[ registar uma conta](https://ers.cr.usgs.gov/login).

This extracted with QGIS SRTM Downloader plugin and clipped)


##### MapBox tiles 
Package ceramic https://github.com/hypertidy/ceramic

https://docs.mapbox.com/help/troubleshooting/access-elevation-data/
Requires an API key
The slopes_3d() function from slopes packages retrieves the z-values information for each vertice, storing an xy linestring as a xyz linestring.


##### Copernicus DEM
The European Land Monitoring Service [@copernicus_2019]  
25m resolution and vertical accuracy of +/- 7m RMSE [@copernicus_2017], for all Europe.


##### IST DEM
This DEM was acquired by Instituto Superior Técnico (University of Lisbon) by 2012, covers all the Northern Metropolitan Area of Lisbon, and has a 10m cell resolution, when projected at the official Portuguese EPSG: 3763 - TM06/ETRS89. No more is known about this raster, and it has been used in several projects at CERIS Research Center.

### For the road network
A sample of Lisbon's Road Network, available on OpenStreetMap.  
After retrieving the data from "portugal" - the only dataset available at the moment for the case study -, we will make a buffer of 2000m around "Campo Martires Patria", right in the center of Lisbon, and collect a sample that contains variability regarding:  

* types of highways, from large avenues to small stairs
* orthogonal and organic highways or streets
* flat and hilly highways
* flat and hilly areas
* long and short highways


## Methods

### To prerare the road netwotk
```{r message=FALSE, warning=FALSE}
#load packages
library(dplyr)
library(sf)
library(osmextract)
library(stplanr)
library(slopes)
library(raster)
library(geodist)
library(tmap)
```

1. Retrieve the OSM road network and filter by highway classes, removing pathways

```{r eval=FALSE}
portugal_osm = oe_get("Portugal", provider = "geofabrik", stringsAsFactors = FALSE, quiet = FALSE, force_download = TRUE, force_vectortranslate = TRUE) #218 MB!
```
```{r eval=FALSE}
portugal_osm_filtered = portugal_osm %>%
  dplyr::filter(
    highway %in% c(
      'primary',"primary_link",'secondary',"secondary_link",
      'tertiary',"tertiary_link","trunk","trunk_link",
      "motorway","motorway_link","service","track",
      "residential","cycleway","living_street","pedestrian",
      "steps", "unclassified"
    )
  )
```
```{r include=FALSE}
# OSMpt = "https://github.com/U-Shift/Declives-RedeViaria/releases/download/0.3/portugal_osm_filtered.Rds"
# portugal_osm_filtered = readRDS(url(u, "rb"))
portugal_osm_filtered = readRDS("D:/GIS/OSM/portugal_osm_filtered.Rds")
```

2. Create a buffer area with 2km around a point in the center of Lisbon, and clip the road network with it

```{r message=FALSE, warning=FALSE}
#buffer area
lisbon_sf = tmaptools::geocode_OSM("campo mártires da pátria", as.sf = TRUE)
lisbon_buffer = stplanr::geo_buffer(shp = lisbon_sf, dist = 2000)
#clip
osm_lines_lisbon = st_crop(portugal_osm_filtered, lisbon_buffer) %>%  #clip by bounding box
                    st_intersection(lisbon_buffer)
```

3. Clean the road netwok, by removing unconnected segments

```{r message=FALSE, warning=FALSE}
osm_lines_lisbon$group = stplanr::rnet_group(osm_lines_lisbon)
plot(osm_lines_lisbon["group"])
osm_lines_lisbon_clean = osm_lines_lisbon %>% filter(group == 1) #keep only the main network cluster

# st_geometry(osm_lines_lisbon_clean) #geometry type:  GEOMETRY || it is required to be LINESTRING
```

4. Filter from the OSM original network, the segments in the clean one

```{r}
RoadNetwork = portugal_osm_filtered %>% filter(osm_id %in% osm_lines_lisbon_clean$osm_id) #ficar apenas os segmentos da rede limpa
```

5. Breaking up the road segments at their internal vertices, but leaving _brunels_ intact

```{r message=TRUE, warning=TRUE}
RoadNetwork = stplanr::rnet_breakup_vertices(RoadNetwork)
nrow(RoadNetwork)
```
```{r eval=FALSE, include=FALSE}
mapview::mapview(RoadNetwork)
```


### To estimate the slope of road network segments
##### With NASA DEM

* Import the DEM and make sure the road network dataset has the same projection

```{r}
demNASA = raster::raster("raster/LisboaNASA_clip.tif")
summary(values(demNASA))
```
```{r echo=FALSE}
raster::plot(demNASA)
plot(sf::st_geometry(RoadNetwork), add = TRUE)
```

* Estimate the gradient

```{r}
RoadNetworkNASA = RoadNetwork
RoadNetworkNASA$slope = slope_raster(RoadNetworkNASA, e = demNASA)
RoadNetworkNASA$slope_pct = RoadNetworkNASA$slope*100 #percentage
```

##### With Map Box

* Estimate the gradient, directly with `slopes`

```{r warning=FALSE}
RoadNetworkMBox = slope_3d(r= RoadNetwork)
RoadNetworkMBox$slope = slope_xyz(RoadNetworkMBox)
RoadNetworkMBox$slope_pct = RoadNetworkMBox$slope*100 #percentage
```

##### With EU-DEM

* Import the DEM and make sure the road network dataset has the same projection

```{r message=FALSE, warning=FALSE}
demEU = raster::raster("raster/LisboaCOPERNICUS_clip.tif")
crs(demEU) = CRS('+init=EPSG:3035') #assign official projection (ETRS89-LAEA)
```
```{r echo=FALSE}
RoadNetworkEU = st_transform(RoadNetwork, 3035) #to the same projection as demEU
raster::plot(demEU)
plot(sf::st_geometry(RoadNetworkEU), add = TRUE)
```

* Estimate the gradient

```{r}
RoadNetworkEU$slope = slope_raster(RoadNetworkEU, e = demEU)
RoadNetworkEU$slope_pct = RoadNetworkEU$slope*100 #percentage
```


##### With IST DEM

* Import the DEM and make sure the road network dataset has the same projection

```{r message=FALSE, warning=FALSE}
demIST = raster::raster("raster/LisboaIST_clip_r1.tif")
crs(demIST) = CRS('+init=EPSG:3763') #assign official projection (Portugal TM06/ETRS89)
```
```{r echo=FALSE}
RoadNetworkIST = st_transform(RoadNetwork, 3763) #to the same projection as demIST
raster::plot(demIST)
plot(sf::st_geometry(RoadNetworkIST), add = TRUE)
```

* Estimate the gradient

```{r}
RoadNetworkIST$slope = slope_raster(RoadNetworkIST, e = demIST)
RoadNetworkIST$slope_pct = RoadNetworkIST$slope*100 #percentage
```


## Results

#### Compare the used DEM values
```{r echo=FALSE, warning=FALSE}
t0 = rbind(summary(values(demNASA)), summary(values(demEU)), summary(values(demIST)))
row.names(t0) = c("STRM NASA", "EU-DEM Copernicus", "IST DEM")
resolution = c(res(demNASA)[1]*100000,res(demEU)[1],res(demIST)[1])
t0 = cbind(resolution, t0)
knitr::kable(t0, digits = 2, caption = "DEM information for the area")
```


#### Compare the estimated gradient values for each method

```{r echo=FALSE}
#filter roads that are common
RoadNetworkEU = RoadNetworkEU %>% filter(!is.na(RoadNetworkEU$slope_pct))
RoadNetworkNASA = RoadNetworkNASA %>% filter(osm_id %in% RoadNetworkEU$osm_id)
RoadNetworkMBox = RoadNetworkMBox %>% filter(osm_id %in% RoadNetworkEU$osm_id)
RoadNetworkIST = RoadNetworkIST %>% filter(osm_id %in% RoadNetworkEU$osm_id)

#make common summary table
t1 = rbind(summary(RoadNetworkNASA$slope_pct), summary(RoadNetworkMBox$slope_pct),
           summary(RoadNetworkEU$slope_pct), summary(RoadNetworkIST$slope_pct))
row.names(t1) = c("STRM NASA", "Ceramic MapBox", "EU-DEM Copernicus", "IST DEM")
knitr::kable(t1, digits = 3, caption = "Result summaries")
```


* Adopt a simplistic qualitative classification for cycling effort uphill, and compare the number of segments in each class

```{r include=FALSE}
RoadNetworkNASA$slope_class = RoadNetworkNASA$slope_pct %>%
  cut(breaks = c(0, 3, 5, 8, 10, 20, Inf),
    labels = c("0-3: flat", "3-5: mild", "5-8: medium", "8-10: hard", "10-20: extreme", ">20: impossible"),
    right = F)
RoadNetworkMBox$slope_class = RoadNetworkMBox$slope_pct %>%
  cut(breaks = c(0, 3, 5, 8, 10, 20, Inf),
    labels = c("0-3: flat", "3-5: mild", "5-8: medium", "8-10: hard", "10-20: extreme", ">20: impossible"),
    right = F)
RoadNetworkEU$slope_class = RoadNetworkEU$slope_pct %>%
  cut(breaks = c(0, 3, 5, 8, 10, 20, Inf),
    labels = c("0-3: flat", "3-5: mild", "5-8: medium", "8-10: hard", "10-20: extreme", ">20: impossible"),
    right = F)
RoadNetworkIST$slope_class = RoadNetworkIST$slope_pct %>%
  cut(breaks = c(0, 3, 5, 8, 10, 20, Inf),
    labels = c("0-3: flat", "3-5: mild", "5-8: medium", "8-10: hard", "10-20: extreme", ">20: impossible"),
    right = F)

```
```{r echo=FALSE}
propNASA = round(prop.table(table(RoadNetworkNASA$slope_class))*100,1)
propMB = round(prop.table(table(RoadNetworkMBox$slope_class))*100,1) 
propEU = round(prop.table(table(RoadNetworkEU$slope_class))*100,1) 
propIST = round(prop.table(table(RoadNetworkIST$slope_class))*100,1) 

t2 = rbind(propNASA, propMB, propEU, propIST)
row.names(t2) = c("STRM NASA", "Ceramic MapBox", "EU-DEM Copernicus", "IST DEM")
knitr::kable(t2, digits = 1, caption = "Percentage of road segments in each gradient interval and qualitative class")
```

* View maps side by side

```{r echo=FALSE, message=FALSE, warning=FALSE}
palredgreen = c("#267300", "#70A800", "#FFAA00", "#E60000", "#A80000", "#730000") #color palette
tmap_mode("plot")
slopesNASA = 
  tm_layout(title = "SRTM 27m", legend.show = F) +
  tm_shape(RoadNetworkNASA) +
  tm_lines(
    col = "slope_class",
    palette = palredgreen, 
    lwd = 1.8, 
    title.col = "Gradient [%]"
  )
slopesMBox = 
  tm_layout(title = "MapBox", legend.show = F) +
  tm_shape(RoadNetworkMBox) +
  tm_lines(
    col = "slope_class",
    palette = palredgreen, 
    lwd = 1.8, 
    title.col = "Gradient [%]"
  )
RoadNetworkEU = st_transform(RoadNetworkEU, 3857) #re-project to be coherent with other fellows
slopesEU = 
  tm_layout(title = "EU-DEM 25m", legend.show = F) +
  tm_shape(RoadNetworkEU) +
  tm_lines(
    col = "slope_class",
    palette = palredgreen, 
    lwd = 1.8, 
    title.col = "Gradient [%]"
  )
slopesIST = 
  tm_layout(title = "IST 10m", legend.show = F) +
  tm_shape(RoadNetworkIST) +
  tm_lines(
    col = "slope_class",
    palette = palredgreen, 
    lwd = 1.8, 
    title.col = "Gradient [%]"
  )
tmap_arrange(slopesNASA, slopesMBox, slopesEU, slopesIST) #map grid
tm_shape(RoadNetworkIST) +
  tm_lines(
    col = "slope_class",
    palette = palredgreen,
    lwd = 1.8,
    title.col = "Gradient [%]"
  ) +
  tm_layout(legend.only = T)
```



## Discussion
_Google Elevation data will also be compared._

## References




